---
title: Rolling functions on non-consecutive grouped data using dplyr and zoo
author: ~
date: '2018-04-26'
slug: dplyr-zoo-rollapply
categories: [R]
tags: [zoo, tidyverse]
output:
  blogdown::html_page:
    # toc: true
    # toc_depth: 1
    number_sections: true
    # code_folding: hide
---

When working with time-series data in R, `zoo` is an indispensable tool to keep on your bench. Here, I  demonstrate a use case for one of the package's best functions, `rollapply`[^1], and one solution for handling non-consecutive time-series data.

[^1]: Allows the user to apply a function via a rolling window over an array.

In our hypothetical experiment, we have three individuals ("A", "B", and "C") who were supposed to have data at ten consecutive time-points. However, multiple time-points for each individual are missing in our final dataset. We want to apply a rolling function to this data and need to ensure that non-consecutive days are treated as such.

An example dataset is displayed below:

```{r, warning=FALSE, message=FALSE}
library(tidyverse)
library(zoo)

dat <- tibble(ID = factor(rep(LETTERS[1:3], each = 10)),
              Timepoint = rep(1:10, 3),
              Value = floor(rnorm(30, 10, 2.5))) %>% 
  filter(!(ID == "A" & Timepoint >= 5), !(ID == "B" & Timepoint %in% c(4,5,6)),
         !(ID == "C" & Timepoint %in% c(3, 8)))
```

```{r, echo=F}
head(dat, 10) %>% 
  knitr::kable(.)
```

To ensure non-consecutive days are treated differently, we'll write a quick function to enumerate the continous data in an array, called `enumerate_cont`.

```{r}
enumerate_cont <- function(x) {
  r <- rle(x - seq_along(x))
  r$values <- seq_along(r$values)
  return(inverse.rle(r))
}
```

As an example:

```{r}
x <- c(1:3, 5:7, 9)
y <- enumerate_cont(x)
names(x) <- y
x
```

We can then use `dplyr::group_by` to apply our `enumerate_cont` function to each indivudal in our dataset. We'll store the enumerated periods in the column `Bout`.

```{r}
dat <- dat %>% 
  group_by(ID) %>% 
  mutate(Bout = enumerate_cont(Timepoint))
```

```{r, echo=F}
head(dat, 10) %>% 
  knitr::kable(.)
```

Finally, we can again use `dplyr::group_by` to apply our rolling function[^2], in this case a moving product, grouping on both the `ID` and `Bout` columns. An `NA` in the `Rolling_Product` column indicates a break in either of our two grouping variables, as specified by the `fill` argument.

[^2]: The other parameters passed to `zoo::rollapply` in this example specify the window size (`width`), the padding/fill value (`fill`), and the window alignment (`align`). 

```{r}
dat <- dat %>% 
  group_by(ID, Bout) %>% 
  mutate(Rolling_Product = zoo::rollapply(data = Value, width = 2,
                                          FUN = function(x) Reduce(`*`, x),
                                          fill = NA, align = "right"))
```

```{r, echo=F}
dat %>% 
  knitr::kable(.)
```

That's it! This is a case I run into constantly, and hope that this gist helps someone else out there.